---
title: "ObjectConversion"
output: html_document
date: "2023-07-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

There are many tools have been developed to process scRNA-seq data, such as [Scanpy](https://scanpy.readthedocs.io/en/stable/), [Seurat](https://satijalab.org/seurat/), [scran](https://bioconductor.org/packages/release/bioc/html/scran.html) and [Monocle](http://cole-trapnell-lab.github.io/monocle-release/). These tools have their own objects, such as `Anndata` of `Scanpy`, `SeuratObject` of `Seurat`, `SingleCellExperiment` of `scran` and `CellDataSet`/`cell_data_set` of `Monocle2`/`Monocle3`. There are also some file format designed for large omics datasets, such as [loom](http://loompy.org/). To perform a comprehensive scRNA-seq data analysis, we usually need to combine multiple tools, which means we need to perform object conversion frequently.

To facilitate user analysis of scRNA-seq data, `GEfetch2R`:

* **benchmarked the format conversion tools** and provides some guides for tool selection under different scenarios(`Anndata` -> `SeuratObject`, `SeuratObject` to `Anndata`, `Anndata` to `SingleCellExperiment`, `SingleCellExperiment` to `Anndata`)
* provides multiple functions to **perform format conversion between widely used scRNA-seq objects** (`SeuratObject`, `AnnData`, `SingleCellExperiment`, `CellDataSet/cell_data_set` and `loom`)

<hr />

## Test data

```{r test_data}
# library
library(GEfetch2R)
library(Seurat) # pbmc_small
# library(scRNAseq) # seger
```

`SeuratObject`:

```{r test_seurat}
# object
pbmc_small
# metadata
head(pbmc_small@meta.data)
```

`SingleCellExperiment`:
```{r testsce}
# seger <- scRNAseq::SegerstolpePancreasData()
# load from local
seger = readRDS("/Users/soyabean/Desktop/tmp/scdown/conversion/seger.rds")
seger
```

<hr />

## Convert SeuratObject to other objects

Here, we will convert SeuratObject to `SingleCellExperiment`, `CellDataSet`/`cell_data_set`, `Anndata`, `loom`.

### SeuratObject to SingleCellExperiment

The conversion is performed with functions implemented in `Seurat`:
```{r seu2sce}
sce.obj <- ExportSeurat(seu.obj = pbmc_small, assay = "RNA", to = "SCE")
sce.obj
```

<hr />

### SeuratObject to CellDataSet/cell_data_set

To `CellDataSet` (The conversion is performed with functions implemented in `Seurat`):

```{r seu2cds1}
# BiocManager::install("monocle") # reuqire monocle
cds.obj <- ExportSeurat(seu.obj = pbmc_small, assay = "RNA", reduction = "tsne", to = "CellDataSet")
cds.obj
```

To `cell_data_set` (The conversion is performed with functions implemented in `SeuratWrappers`):

```{r seu2cds2, eval=FALSE}
# remotes::install_github('cole-trapnell-lab/monocle3') # reuqire monocle3
cds3.obj <- ExportSeurat(seu.obj = pbmc_small, assay = "RNA", to = "cell_data_set")
cds3.obj
```

<hr />

### SeuratObject to AnnData

There are multiple tools available for format conversion from `SeuratObject` to `Anndata`:

* `scDIOR` is the best method in terms of information kept and usability
* `sceasy` has best performance in running time and disk usage.

```{r seu2anndata}
# sceasy
Seu2AD(
  seu.obj = pbmc_small, method = "sceasy", out.folder = "/Users/soyabean/Desktop/tmp/scdown/conversion",
  assay = "RNA", slot = "counts", conda.path = "/Applications/anaconda3"
)
# # SeuratDisk
# Seu2AD(seu.obj = pbmc_small, method = "SeuratDisk", out.folder = "/Users/soyabean/Desktop/tmp/scdown/conversion",
#        assay = "RNA", save.scale = TRUE)
# # scDIOR
# Seu2AD(seu.obj = pbmc_small, method = "scDIOR",
#        out.folder = "/Users/soyabean/Desktop/tmp/scdown/conversion", assay = "RNA", save.scale = TRUE)
```

<hr />

### SeuratObject to loom

The conversion is performed with functions implemented in `SeuratDisk`:
```{r seu2loom}
ExportSeurat(
  seu.obj = pbmc_small, assay = "RNA", to = "loom",
  loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small.loom"
)
```

<hr />

## Convert other objects to SeuratObject

### SingleCellExperiment to SeuratObject

The conversion is performed with functions implemented in `Seurat`:
```{r sce2seu}
seu.obj.sce <- ImportSeurat(
  obj = sce.obj, from = "SCE", count.assay = "counts",
  data.assay = "logcounts", assay = "RNA"
)
seu.obj.sce
```

<hr />

### CellDataSet/cell_data_set to SeuratObject

`CellDataSet` to `SeuratObject` (The conversion is performed with functions implemented in `Seurat`):
```{r cds2seu1}
seu.obj.cds <- ImportSeurat(
  obj = cds.obj, from = "CellDataSet",
  count.assay = "counts", assay = "RNA"
)
seu.obj.cds
```

`cell_data_set` to `SeuratObject` (The conversion is performed with functions implemented in `Seurat`):
```{r cds2seu2, eval=FALSE}
seu.obj.cds3 <- ImportSeurat(
  obj = cds3.obj, from = "cell_data_set",
  count.assay = "counts", data.assay = "logcounts", assay = "RNA"
)
seu.obj.cds3
```

<hr />

### AnnData to SeuratObject

There are multiple tools available for format conversion from `AnnData` to `SeuratObject`:

* `scDIOR` is the best method in terms of information kept (**`GEfetch2R` integrates `scDIOR` and `SeuratDisk` to achieve the best performance in information kept**)
* `schard` is the best method in terms of usability
* `schard` and `sceasy` have comparable performance when cell number below 200k, but `sceasy` has better performance in scalability
* `sceasy` has better performance in disk usage 

```{r anndata2seu}
# sceasy
ann.sceasy <- AD2Seu(
  anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small_sceasy.h5ad", method = "sceasy",
  assay = "RNA", slot = "scale.data"
)
ann.sceasy
# # SeuratDisk
# ann.seu <- AD2Seu(anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small_sceasy.h5ad",
#                   method = "SeuratDisk", assay = "RNA", load.assays = c("RNA"))
# # scDIOR
# ann.scdior <- AD2Seu(anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small_sceasy.h5ad",
#                      method = "scDIOR", assay = "RNA")
# # schard
# ann.schard <- AD2Seu(anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small_sceasy.h5ad",
#                      method = "schard", assay = "RNA", use.raw = T)
# # SeuratDisk+scDIOR
# ann.seuscdior <- AD2Seu(anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small_sceasy.h5ad",
#                         method = "SeuratDisk+scDIOR", assay = "RNA", load.assays = c("RNA"))
```

<hr />

### loom to SeuratObject

The conversion is performed with functions implemented in `SeuratDisk` and `Seurat`:

```{r loom2seu}
# loom will lose reduction
seu.obj.loom <- ImportSeurat(loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small.loom", from = "loom")
seu.obj.loom
```

<hr />

## Conversion between SingleCellExperiment and AnnData

### SingleCellExperiment to AnnData

There are multiple tools available for format conversion from `SingleCellExperiment` to `AnnData`:

* `zellkonverter` is the best method in terms of information kept and running time
* `scDIOR` is the best method in terms of usability and disk usage

```{r sce2anndata}
# zellkonverter
SCE2AD(
  sce.obj = seger, method = "zellkonverter",
  out.folder = "/Users/soyabean/Desktop/tmp/scdown/conversion", slot = "counts",
  conda.path = "/Applications/anaconda3"
)

# # sceasy
# SCE2AD(sce.obj = seger, method = "sceasy", out.folder = "/Users/soyabean/Desktop/tmp/scdown/conversion",
#        slot = "counts", conda.path = "/Applications/anaconda3")
#
# # scDIOR
# seger.scdior <- seger
# library(SingleCellExperiment)
# # scDIOR does not support varm in rowData
# rowData(seger.scdior)$varm <- NULL
# SCE2AD(sce.obj = seger.scdior, method = "scDIOR", out.folder = "/Users/soyabean/Desktop/tmp/scdown/conversion")
```

<hr />

### AnnData to SingleCellExperiment

There are multiple tools available for format conversion from `AnnData` to `SingleCellExperiment`:

* `zellkonverter` is the best method in terms of information kept
* `schard` is the best method in terms of usability and running time
* `schard` and `scDIOR` have comparable performance in disk usage

```{r anndata2sce}
# zellkonverter
sce.zell <- AD2SCE(
  anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger_zellkonverter.h5ad",
  method = "zellkonverter", slot = "scale.data",
  use.raw = TRUE, conda.path = "/Applications/anaconda3"
)
sce.zell

# # scDIOR
# sce.scdior <- AD2SCE(anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger_zellkonverter.h5ad.h5ad",
#                      method = "scDIOR", assay = "RNA",
#                      use.raw = TRUE, conda.path = "/Applications/anaconda3")
# # schard
# sce.schard <- AD2SCE(anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger_zellkonverter.h5ad.h5ad",
#                      method = "schard", use.raw = TRUE)
```

<hr />

## Conversion between SingleCellExperiment and loom

The conversion is performed with functions implemented in `LoomExperiment`.

### SingleCellExperiment to loom

```{r sce2loom, eval=FALSE}
# remove seger.loom first
SCELoom(
  from = "SingleCellExperiment", to = "loom", sce = seger,
  loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger.loom"
)
```

<hr />

### loom to SingleCellExperiment

```{r loom2sce}
seger.loom <- SCELoom(
  from = "loom", to = "SingleCellExperiment",
  loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger.loom"
)
seger.loom
```

<hr />













