---
title: "ObjectConversion"
output: html_document
date: "2023-07-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

There are many tools have been developed to process scRNA-seq data, such as [Scanpy](https://scanpy.readthedocs.io/en/stable/), [Seurat](https://satijalab.org/seurat/), [scran](https://bioconductor.org/packages/release/bioc/html/scran.html) and [Monocle](http://cole-trapnell-lab.github.io/monocle-release/). These tools have their own objects, such as `Anndata` of `Scanpy`, `SeuratObject` of `Seurat`, `SingleCellExperiment` of `scran` and `CellDataSet`/`cell_data_set` of `Monocle2`/`Monocle3`. There are also some file format designed for large omics datasets, such as [loom](http://loompy.org/). To perform a comprehensive scRNA-seq data analysis, we usually need to combine multiple tools, which means we need to perform object conversion frequently. To facilitate user analysis of scRNA-seq data, `scfetch` provides multiple functions to perform object conversion between widely used tools and formats. The object conversion implemented in `scfetch` has two main advantages: 

* **one-step conversion between different objects**. There will be no conversion to intermediate objects, thus preventing unnecessary information loss.
* **tools used for object conversion are developed by the team of the source/destination object as far as possible**. For example, we use `SeuratDisk` to convert SeuratObject to loom, use `zellkonverter` to perform conversion between `SingleCellExperiment` and `Anndata`. When there is no such tools, we use `sceasy` to perform conversion.

<hr />

## Test data

```{r test_data}
# library
library(scfetch)
library(Seurat) # pbmc_small
library(scRNAseq) # seger
```

`SeuratObject`:

```{r test_seurat}
# object
pbmc_small
# metadata
head(pbmc_small@meta.data)
```

`SingleCellExperiment`:
```{r testsce}
seger <- scRNAseq::SegerstolpePancreasData()
seger
```

<hr />

## Convert SeuratObject to other objects

Here, we will convert SeuratObject to `SingleCellExperiment`, `CellDataSet`/`cell_data_set`, `Anndata`, `loom`.

### SeuratObject to SingleCellExperiment

The conversion is performed with functions implemented in `Seurat`:
```{r seu2sce}
sce.obj <- ExportSeurat(seu.obj = pbmc_small, assay = "RNA", to = "SCE")
sce.obj
```

<hr />

### SeuratObject to CellDataSet/cell_data_set

To `CellDataSet` (The conversion is performed with functions implemented in `Seurat`):

```{r seu2cds1}
# BiocManager::install("monocle") # reuqire monocle
cds.obj <- ExportSeurat(seu.obj = pbmc_small, assay = "RNA", reduction = "tsne", to = "CellDataSet")
cds.obj
```

To `cell_data_set` (The conversion is performed with functions implemented in `SeuratWrappers`):

```{r seu2cds2}
# remotes::install_github('cole-trapnell-lab/monocle3') # reuqire monocle3
cds3.obj <- ExportSeurat(seu.obj = pbmc_small, assay = "RNA", to = "cell_data_set")
cds3.obj
```

<hr />

### SeuratObject to AnnData

`AnnData` is a Python object, `reticulate` is used to communicate between Python and R. User should create a Python environment which contains `anndata` package and specify the environment path with `conda.path` to ensure the exact usage of this environment.

The conversion is performed with functions implemented in `sceasy`:
```{r seu2anndata, eval=FALSE}
# remove pbmc_small.h5ad first
ExportSeurat(
  seu.obj = pbmc_small, assay = "RNA", to = "AnnData", conda.path = "/Applications/anaconda3",
  anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small.h5ad"
)
```

<hr />

### SeuratObject to loom

The conversion is performed with functions implemented in `SeuratDisk`:
```{r seu2loom, eval=FALSE}
ExportSeurat(
  seu.obj = pbmc_small, assay = "RNA", to = "loom",
  loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small.loom"
)
```

<hr />

## Convert other objects to SeuratObject

### SingleCellExperiment to SeuratObject

The conversion is performed with functions implemented in `Seurat`:
```{r sce2seu}
seu.obj.sce <- ImportSeurat(obj = sce.obj, from = "SCE", count.assay = "counts", data.assay = "logcounts", assay = "RNA")
seu.obj.sce
```

<hr />

### CellDataSet/cell_data_set to SeuratObject

`CellDataSet` to `SeuratObject` (The conversion is performed with functions implemented in `Seurat`):
```{r cds2seu1}
seu.obj.cds <- ImportSeurat(obj = cds.obj, from = "CellDataSet", count.assay = "counts", assay = "RNA")
seu.obj.cds
```

`cell_data_set` to `SeuratObject` (The conversion is performed with functions implemented in `Seurat`):
```{r cds2seu2}
seu.obj.cds3 <- ImportSeurat(obj = cds3.obj, from = "cell_data_set", count.assay = "counts", data.assay = "logcounts", assay = "RNA")
seu.obj.cds3
```

<hr />

### AnnData to SeuratObject

`AnnData` is a Python object, `reticulate` is used to communicate between Python and R. User should create a Python environment which contains `anndata` package and specify the environment path with `conda.path` to ensure the exact usage of this environment.

The conversion is performed with functions implemented in `sceasy`:
```{r anndata2seu}
seu.obj.h5ad <- ImportSeurat(
  anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small.h5ad",
  from = "AnnData", assay = "RNA", conda.path = "/Applications/anaconda3"
)
seu.obj.h5ad
```

<hr />

### loom to SeuratObject

The conversion is performed with functions implemented in `SeuratDisk` and `Seurat`:

```{r loom2seu}
# loom will lose reduction
seu.obj.loom <- ImportSeurat(loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/pbmc_small.loom", from = "loom")
seu.obj.loom
```

<hr />

## Conversion between SingleCellExperiment and AnnData

The conversion is performed with functions implemented in `zellkonverter`.

### SingleCellExperiment to AnnData

```{r sce2anndata, eval=FALSE}
# remove seger.h5ad first
SCEAnnData(
  from = "SingleCellExperiment", to = "AnnData", sce = seger, X_name = "counts",
  anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger.h5ad"
)
```

<hr />

### AnnData to SingleCellExperiment

```{r anndata2sce}
seger.anndata <- SCEAnnData(
  from = "AnnData", to = "SingleCellExperiment",
  anndata.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger.h5ad"
)
seger.anndata
```

<hr />

## Conversion between SingleCellExperiment and loom

The conversion is performed with functions implemented in `LoomExperiment`.

### SingleCellExperiment to loom

```{r sce2loom, eval=FALSE}
# remove seger.loom first
SCELoom(
  from = "SingleCellExperiment", to = "loom", sce = seger,
  loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger.loom"
)
```

<hr />

### loom to SingleCellExperiment

```{r loom2sce}
seger.loom <- SCELoom(
  from = "loom", to = "SingleCellExperiment",
  loom.file = "/Users/soyabean/Desktop/tmp/scdown/conversion/seger.loom"
)
seger.loom
```

<hr />













